---
title: "dnds_M0_compare_branches_methods"
author: "KN"
date: "2/24/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup2}

library(lme4)
library(sjPlot)
library(stargazer)
library(ggplot2)
library(viridis)
library(tidyverse)
library(gridExtra)
library(ggpubr)
library(nord)

TXT_SIZE=14
LINE_SIZE=0.2

dnds_colour <- "#778868"
dnds_gradient_col <- c("#b5b991", "#778868","#3d5941")


sessionInfo()
```

## Result codeml M0 and branch models

#Model M0 
Average dnds per gene across branches

Distribution of dnds in the aligned geneset using codeml Model M0. Average dnds per gene across branches.
```{r codeml M0, echo=FALSE}

#loading data from codeml M0 unfiltered and filtered
Result_M0_r1 <- read.table("data/Model_M0/summary_results_M0run1.table", header = T)

#filtered dS< 30
Result_M0_r1_filtered <- read.table("data/Model_M0/summary_results_M0run1_filtered.table", header = T)

#run2
Result_M0_r2 <- read.table("data/Model_M0/summary_results_M0_run2.table", header = T)

#filtered dS< 30
Result_M0_r2_filtered <- read.table("data/Model_M0/summary_results_M0run2_filtered.table", header = T)

###############
#####checking correlation between runs
cor.test(Result_M0_r1$Omega.dN.dS., Result_M0_r2$Omega.dN.dS.)


```

Very high correlation between runs, using only run1 that is more conservative resulting in 4237 genes (run 2: 4239 genes)

```{r M0_result}

median(Result_M0_r1_filtered$Omega.dN.dS.) #[1] 0.04037
mean(Result_M0_r1_filtered$Omega.dN.dS.) #[1] 0.04451873
sd(Result_M0_r1_filtered$Omega.dN.dS.) #[1] 0.02545054

capture.output(summary_M0, file = "Result_tables/summary_M0.txt")
#add sd manually

```

```{r M0_plot, echo=FALSE}


#graphs of filtered M0
hist(Result_M0_r1$Tree_len_dS, col = "Brown", main = "Tree length for dS", breaks = 100)
hist(Result_M0_r1_filtered$Tree_len_dS, col = "Brown", main = "Tree length for dS", breaks = 100)

TXT_SIZE=14
LINE_SIZE=0.2
fall_colours <- carto_pal(n=7, "Fall")
#colour_dnds <- "#32324BDD"
dnds_colour <- "#778868"
dnds_gradient_col <- c("#b5b991", "#778868","#3d5941")

ggplot(Result_M0_r1_filtered, aes(Omega.dN.dS.)) + 
  geom_density(colour="black", fill=dnds_colour, size = 0.2) +
  labs(x=expression(omega),
       y="Density") +
  theme(axis.text.x = element_text(vjust = 0.5,
                                   size = TXT_SIZE, colour = "black"),
        axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), 
                                    size = TXT_SIZE),
        axis.text.y = element_text(size = TXT_SIZE, colour = "black"),
        axis.title.y = element_text(size = TXT_SIZE, 
                                    margin = margin(t = 0, r = 10, b = 0, l = 0)),
        panel.background = element_blank(), 
        axis.line = element_line(size = LINE_SIZE),
        axis.ticks = element_line(size = LINE_SIZE),
        axis.ticks.length = unit(2, "pt"),
        plot.margin = margin(t = 20, r = 20, b = 20, l = 20, unit = "pt")) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0))

ggsave("Figure_ms/average_w_green.pdf", device = "pdf", width = 8, height = 6)

```


# Evolutionary rate per branch (codeml with removing gaps (clean=1))
filtering filter out rows where any value is below/above the threshold 
dnds < 10
 If dS is less than 0,01 the dnds estimation is unreliable
 If dS is > 2 risk of saturation, although some reports that codeml is rather stable to higher dS so filtering on total dS across branches 30

```{r branch_data_clean, echo=FALSE}
branch_dnds_tip <- read.table("data/result_branch_dnds_clean_tip.csv", header = T)

#dS
branch_ds_tip <- read.table("data/result_branch_ds_clean_tip.csv", header = T)

#filtering
#where the total tree length is above 30
length(branch_ds_tip[!rowSums(branch_ds_tip[2:9]) > 30,2])
#3060
branch_ds_tip[rowSums(branch_ds_tip[2:9]) > 30,]

branch_ds_tip_filtr30 <- branch_ds_tip[!rowSums(branch_ds_tip[2:9]) > 30,]

hist(rowSums(branch_ds_tip[2:9]), breaks = 1000, xlim = c(0,500))
hist(rowSums(branch_ds_tip_filtr30[2:9]), breaks = 500, xlim = c(0,1))

hist(branch_ds_tip[,2], breaks = 2000, xlim = c(0,5))

#filter dnds
branch_dnds_filtr30 <- branch_dnds_tip[branch_dnds_tip$GeneID %in% branch_ds_tip_filtr30$GeneID, ]

#filter dnds < 10 across branches
branch_dnds_filtr30_10 <- branch_dnds_filtr30[rowSums(branch_dnds_filtr30[2:9]) < 10,]

#filter dN
branch_dn_tip <- read.table("data/result_branch_dn_clean_tip.csv", header = T)

branch_dn_filtr30 <- branch_dn_tip[branch_dn_tip$GeneID %in% branch_ds_tip_filtr30$GeneID, ]

#long
#data_long <- gather(olddata_wide, condition, measurement, control:cond2, factor_key=TRUE)
branch_dnds_long <- gather(branch_dnds_filtr30_10, key = branch, value = dnds, -GeneID)

branch_ds_long <- gather(branch_ds_tip_filtr30, key = branch, value = ds, -GeneID)

branch_dn_long <- gather(branch_dn_filtr30, key = branch, value = dn, -GeneID)

branch_paml_long <- inner_join(inner_join(branch_dn_long, branch_ds_long), branch_dnds_long)

branch_paml_long <- branch_paml_long[!is.na(branch_paml_long$dn), ]

#filter ds < 0.01
branch_paml_long <- branch_paml_long[branch_paml_long$ds > 0.01, ]


length(unique(branch_paml_long$GeneID))

#keep only genes present in all species
gene_list <- 
branch_paml_long %>%
dplyr::count(GeneID) %>%
  filter(n==8) %>%
  select(GeneID)

branch_paml_long <- branch_paml_long[branch_paml_long$GeneID %in% gene_list$GeneID,]

branch_paml_long$branch <- as.factor(branch_paml_long$branch)
branch_paml_long$GeneID <- as.factor(branch_paml_long$GeneID)

length(unique(branch_paml_long$GeneID))
#2696 genes

branch_paml_long$branch <- as.factor(branch_paml_long$branch)
#change order
levels(branch_paml_long$branch)
#[1] "B_mori"      "C_cecrops"   "D_plexippus" "H_melpomene" "L_accius"    "L_sinapis"   "P_machaon"   "P_sennae"   
#reassign
levels(branch_paml_long$branch) <- c("B. mori", "C. cecrops", "D. plexippus", "H. melpomene", "L. accius", "L. sinapis", "P. machaon", "P. sennae")

#reorder
branch_paml_long$branch <- factor(branch_paml_long$branch, levels=c("B. mori","P. machaon", "L. accius", "L. sinapis", "P. sennae", "C. cecrops", "D. plexippus", "H. melpomene"))

summary_branch <- 
branch_paml_long %>%
  group_by(branch) %>%
  dplyr::summarise(dnds=mean(dnds_paml), dnds_sd=sd(dnds_paml), dn=mean(dn_paml), dn_sd=sd(dn_paml), ds=mean(ds_paml), ds_sd=sd(ds_paml))
    #mutate(across(-branch, num, digits = 3, notation = "sci"))

subst_type_gccons_tot <- read.table(file = "../08_subst_type/08_03_subst_type_R/data/subst_type_gccons_tot_gccontent.table", sep = " ")

summary_branch_mapNH <- 
subst_type_gccons_tot %>%
  filter(type %in% "all") %>%
  group_by(branch) %>%
  dplyr::summarise(dnds_mapNH=mean(dnds), dnds_sd_mapNH=sd(dnds), dn_mapNH=mean(dn), dn_sd_mapNH=sd(dn), ds_mapNH=mean(ds), ds_sd_mapNH=sd(ds)) 

tab_df(as.data.frame(summary_branch), file = "Result_tables/summary_branch_paml.html", digits = 6)
tab_df(as.data.frame(summary_branch_mapNH), file = "Result_tables/summary_branch_mapNH.html", digits = 6)

#manually improv in xl




```

#Analysing the branch data
Are there any differences in evolutionary rate between the branches?

dnds is a continuous variable with distribution 0 - > ∞
This suggests a gamma distribution (the distribution of the dnds-value is also respresenting a gamma distr. )

```{r test_diff_paml}



mod_gamma <- glmer(dnds~branch + (1|GeneID), data = branch_paml_long, family = Gamma(link = "log"))
glmer(dnds~branch + (1|GeneID), data = branch_paml_long[branch_paml_long$branch!="B_mori",], family = Gamma(link = "log"))

plot_models(mod_gamma, mod_mapnh_dnds) + theme_pubr()

mod_gene_gamma <- glmer(dnds~(1|GeneID), data = branch_paml_long, family = Gamma(link = "log"))

fixef(mod_gamma)
plot(mod_gamma)
qqplot(residuals(mod_gamma))


mod_gamma_dn <- glmer(dn~branch + (1|GeneID), data = branch_paml_long, family = Gamma(link = "log"))

mod_gamma_ds <- glmer(ds~branch + (1|GeneID), data = branch_paml_long, family = Gamma(link = "log"))

tab_model(mod_gamma, mod_gamma_dn, mod_gamma_ds, 
          collapse.ci = TRUE, 
          p.style     = "numeric_stars")

tab_model(mod_gamma)
tab_model(glmer(dnds~branch + (1|GeneID), data = branch_paml_long[branch_paml_long$branch!="B. mori",], family = Gamma(link = "log")))


aggregate(dnds~branch, mean, data = branch_paml_long)


#check mapNH

subst_type_gccons_tot$branch <- as.factor(subst_type_gccons_tot$branch)
#change order
levels(subst_type_gccons_tot$branch)
#[1] "B_mori"      "C_cecrops"   "D_plexippus" "H_melpomene" "L_accius"    "L_sinapis"   "P_machaon"   "P_sennae"   
#reassign
levels(subst_type_gccons_tot$branch) <- c("B. mori", "C. cecrops", "D. plexippus", "H. melpomene", "L. accius", "L. sinapis", "P. machaon", "P. sennae")

#reorder
subst_type_gccons_tot$branch <- factor(subst_type_gccons_tot$branch, levels=c("B. mori","P. machaon", "L. accius", "L. sinapis", "P. sennae", "C. cecrops", "D. plexippus", "H. melpomene"))

#same genes
subst_type_gccons_tot_filt <- subst_type_gccons_tot[subst_type_gccons_tot$gene_ID %in% branch_paml_long$GeneID, ]

#difference between branches
mod_mapnh_dnds <- glmer(dnds~branch + (1|gene_ID), data = subst_type_gccons_tot_filt[subst_type_gccons_tot_filt$type == "all", ], family = Gamma(link = "log"))
mod_mapnh_dnds_unfilt <- glmer(dnds~branch + (1|gene_ID), data = subst_type_gccons_tot[subst_type_gccons_tot$type == "all", ], family = Gamma(link = "log"))

tab_model(mod_mapnh_dnds)
tab_model(mod_mapnh_dnds_unfilt)

tab_model(mod_gamma, mod_mapnh_dnds, mod_mapnh_dnds_unfilt, file = "Result_tables/dnds_comp_branches.doc")

plot_models(mod_mapnh_dnds, mod_mapnh_dnds_unfilt)



mod_dnds <- plot_models(mod_gamma, mod_mapnh_dnds, show.values = TRUE, m.labels = c("PaML, branch-model", "MapNH, branch-model" ), colors = c("sandybrown", "sienna4"))  +
  scale_y_continuous(limits = c(0,2)) +
    theme(panel.background = element_rect(fill="white"),
          axis.line = element_line(colour = "black", size=LINE_SIZE),
          axis.text.y = element_text(size = TXT_SIZE, colour = "black", face="italic"),
          axis.text.x = element_text(size = TXT_SIZE, colour = "black"),
          axis.ticks.x = element_line(size = LINE_SIZE, colour = "black"),
          axis.ticks.y = element_line(size = LINE_SIZE, colour = "black"),
          legend.text = element_text(size = TXT_SIZE),
          legend.title = element_blank(),
          legend.key = element_blank())

#to set the position of the labels (value.offset does not work)
mod_dnds$layers[[4]]$aes_params$vjust<-(2)
mod_dnds$layers[[4]]$aes_params$hjust<-(0.2)



```


```{r plots_branch, echo=FALSE}

#plot distribution dnds per lineage
#add variable with difference in mean between groups
branch_dnds_long$diff_group <- "Low"
branch_dnds_long[branch_dnds_long$branch=="B_mori" | branch_dnds_long$branch=="P_machaon"|branch_dnds_long$branch== "L_accius", ]$diff_group <- "Medium"
branch_dnds_long[branch_dnds_long$branch=="L_sinapis" | branch_dnds_long$branch=="C_cecrops", ]$diff_group <- "High"

branches_dnds <- 
ggplot(branch_dnds_long, 
       aes(x = branch, y = dnds)) +
  #geom_violin(aes(fill=diff_group)) +
  geom_boxplot(aes(fill=diff_group),
               #width = 0.3,
               #notch = TRUE, 
               lwd = 0.4,
               #fatten = 2,
               outlier.colour = "#8C888B",
               outlier.size = 0.4) +
               #show.legend = FALSE) +
  ylab(label = expression(log[10]*" ("~omega~")")) +
  labs(fill = expression("Average "~omega)) +
  scale_y_continuous(trans = "log10",
                     n.breaks = 6) +
  scale_x_discrete(limits = c("B_mori", "P_machaon","L_accius", "L_sinapis", "P_sennae", "C_cecrops", "D_plexippus", "H_melpomene"), 
                   labels = c("L_accius" = "L. accius", "P_sennae" = "P. sennae", "L_sinapis" = "L. sinapis", "C_cecrops" = "C. cecrops", "D_plexippus" = "D. plexippus", "H_melpomene" = "H. melpomene", "P_machaon" = "P. machaon", "B_mori" = "B. mori")) +
  scale_colour_manual(values = "#8C888B",
                      guide = FALSE) +
  scale_fill_manual(limits = c("High", "Medium","Low"),
                    values = dnds_gradient_col) +
  theme(panel.background = element_blank(),
        axis.line = element_line(size = LINE_SIZE, colour = "black"),
        axis.ticks = element_line(size = LINE_SIZE, colour = "black"),
        axis.text.x = element_text(angle = 45, size = TXT_SIZE, face = "italic", hjust = 1, colour = "black"),
        axis.title.x = element_blank(),
        axis.text.y = element_text(size = TXT_SIZE, colour = "black"),
        axis.title.y = element_text(size = TXT_SIZE, colour = "black"),
        legend.text = element_text(size = TXT_SIZE, colour = "black"),
        legend.title = element_text(size = TXT_SIZE, colour = "black"),
        legend.key = element_blank()  )

ggsave("Figure_ms/dnds_branches_box_green.pdf", device = "pdf", width = 8, height = 5)  
ggsave("Figure_ms/dnds_branches_box_green.jpg", device = "pdf", width = 8, height = 5)  

ggsave(
  ggarrange(mod_dnds, branches_dnds, labels = "auto"),
  device = "pdf",
  height = 6, width = 18,
  filename = "Figure_ms/Suppl_2_diff_br.pdf")
  

```

```{r corr_paml_mapnh}

colnames(branch_paml_long) <- c("gene_ID", "branch", "dn_paml", "ds_paml", "dnds_paml")

paml_mapnh <- inner_join(subst_type_gccons_tot, branch_paml_long)

paml_mapnh$branch <- as.factor(paml_mapnh$branch)


length(unique(paml_mapnh[,1]))
length(unique(subst_type_gccons_tot[,1]))


paml_mapnh %>%
  filter(type %in% "all") %>%
ggplot(aes(ds_paml, ds)) +
  geom_point() +
  stat_cor(method = "spearman") +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(limits = c(0,5)) +
  facet_wrap(~branch)

paml_mapnh %>%
  filter(type %in% "all") %>%
ggplot(aes(dn_paml, dn)) +
  geom_point() +
  stat_cor(method = "spearman") +
  scale_y_continuous(limits = c(0,0.8)) +
  scale_x_continuous(limits = c(0,0.8)) +
  facet_wrap(~branch)
  
paml_mapnh %>%
  filter(type %in% "all") %>%
ggplot() +
  geom_histogram(aes(dn/dn_paml), binwidth = 0.1) +
  scale_x_continuous(limits = c(0,5)) +
  facet_wrap(~branch)

paml_mapnh %>%
  filter(type %in% "all") %>%
ggplot() +
  geom_histogram(aes(ds_paml/ds), binwidth = 0.1) +
  scale_x_continuous(limits = c(0,5)) +
  facet_wrap(~branch)

paml_mapnh %>%
  filter(type %in% "all") %>%
ggplot(aes(dnds_paml, dnds)) +
  geom_point() +
  stat_cor(method = "spearman") +
  scale_y_continuous(limits = c(0,0.8)) +
  scale_x_continuous(limits = c(0,0.8)) +
  facet_wrap(~branch)




```